---
title: "NHANES data analysis"
output: html_document
date: "2025-04-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
NHANESresearch <- read.csv("C:/Users/ekatm/OneDrive/Desktop/NHANES Data/data/clean_data")
```


## R Markdown
DATA ANALYSIS OF MULTIVITAMIN USAGE AND HIGH-SENSITIVITY C-PROTEIN AMONG DIFFERENT DEMOGRAPHICS.
```{r}
library(tidyr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(modelsummary)
library(questionr)
library(tidyselect)
```

```{r}
NHANESresearchreduced <- NHANESresearch %>% 
  filter(NHANESresearch$HouseholdIncome== c("<$5,000", "5-10,000", "$10-15,000", "$15-20,000", "$20-25,000", "$25-35,000", "$35-45,000", "$45-55,000", "$55-65,000", "$65-75,000", "$75-100,000", ">$100,000"))

NHANESresearchreduced



```


```{r}
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==1] <-"<$5,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==2] <- "$5-10,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==3] <- "$10-15,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==4] <- "$15-20,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==5] <- "$20-25,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==6] <- "$25-35,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==7] <- "$35-45,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==8] <- "$45-55,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==9] <- "$55-65,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==10] <- "$65-75,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==14] <- "$75-100,000"
NHANESresearch$HouseholdIncome[NHANESresearch$HouseholdIncome==15] <- ">$100,000"


```


```{r}
head(NHANESresearch)
```

#Descriptive Statistics for our variables:

Age
```{r}
NHANESresearch %>% 
  summarize(mean(Age), sd(Age), count=n())


NHANESresearch %>% 
  ggplot(aes(x=Age)) + geom_histogram(bins=35, color="coral", fill="ivory",) + labs(title="Respondent Age", x="Age (Years)")
```
Gender
```{r}
#freq(Var, cum=TRUE, na.omit(Var))
freq(NHANESresearch$Gender, cum=TRUE)
  
NHANESresearch %>%
  ggplot(aes(x=Gender)) + geom_bar(postition="dodge", color="coral", fill="ivory") + labs(title="Respondent Gender")
```

BMI cat
```{r}
freq(NHANESresearch$BMIcat, cum=TRUE)


NHANESresearch %>% 
  drop_na(BMIcat) %>% 
  ggplot(aes(x=BMIcat)) + geom_bar(postition="dodge", color="coral", fill="ivory") + labs(title= "BMI Category", x="BMI category")
```

BMI
```{r}
NHANESresearch %>% 
  drop_na(BMI) %>% 
  summarize(mean(BMI), sd(BMI), count=n())



NHANESresearch %>%
  drop_na(BMI) %>% 
  ggplot(aes(x=BMI)) + geom_histogram(bins=35, color="coral", fill="ivory",) + labs(title="BMI")

```

High-Sensitivity C-Reactive

```{r}
summary(NHANESresearch$HSCReactive)

#IQR= 3.56-0.55=3.01- upper limit is 8.9; everything below 1stQ is acceptable data.

1.5*3.56
.55-5.34 #lowerbound -4.79
3.56+5.34 #upperbound 8.9

NHANESresearch %>% 
  drop_na(HSCReactive) %>% 
  summarize(mean(HSCReactive), sd(HSCReactive), count=n())

NHANESresearch %>% 
  drop_na(HSCReactive) %>% 
  ggplot(aes(x=HSCReactive)) + geom_histogram(color="coral", fill="ivory") + labs(title="High-Sensitivity C-Reactive Protein Levels", x="HSCRP (mg/L)")

#filtered to exclude outliers.

NHANESresearch %>% 
  drop_na(HSCReactive) %>% 
  filter(HSCReactive<8.9) %>% 
  ggplot(aes(x=HSCReactive)) + geom_histogram(color="coral", fill="ivory") + labs(title="High-Sensitivity C-Reactive Protein Levels (Outliers Removed)", x="HSCRP (mg/L)")
```
```{r}
NHANESresearch <- NHANESresearch %>% 
  mutate(HSCRPlevel = case_when(
    HSCReactive >=3 ~"High",
    HSCReactive <3 ~"Healthy"
  ))
NHANESresearch <-NHANESresearch %>% 
  drop_na(HSCRPlevel)

freq(NHANESresearch$HSCRPlevel, cum=TRUE)

NHANESresearch %>% 
  ggplot(aes(x=HSCRPlevel)) + geom_bar(postition="dodge", color="coral", fill="ivory") + labs(title="High Sensitivity C-Reactive Protein Level", subtitle= "Healthy HSCRP <=3mg/L, High HSCRP> 3mg/L")
```


Household Income
```{r}
freq(NHANESresearchreduced$HouseholdIncome, cum=TRUE)


NHANESresearchreduced %>% 
  drop_na(HouseholdIncome) %>% 
  summarize(mean(HouseholdIncome), sd(HouseholdIncome), count=n())


NHANESresearchreduced %>% 
  drop_na(HouseholdIncome) %>% 
  ggplot(aes(x=HouseholdIncome)) + geom_bar(color="coral", fill="ivory") + labs(title="Annual Household Income", x="Annual Household Income (USD)") + theme(axis.text.x = element_text(angle=45, hjust=1, size=10))
```

Multivitamin 
```{r}
freq(NHANESresearch$Multivitamin, cum=TRUE)

NHANESresearch %>% 
  drop_na(Multivitamin) %>% 
  ggplot(aes(x=Multivitamin)) + geom_bar(postition="dodge", color="coral", fill="ivory") + labs(title="Daily Multivitamin Supplement", x="'Do You Take a Daily Dietary Supplement?'")


```


#Bivariate: Who's taking their vitamins?
```{r}

income_multivitamin <- xtabs(~Multivitamin + HouseholdIncome, data=NHANESresearch)
prop.table(income_multivitamin)*100

NHANESresearchreduced %>% 
  drop_na(Multivitamin) %>% 
  ggplot(aes(x=Multivitamin)) + geom_bar(postition="dodge", color="coral", fill="ivory") + facet_wrap(~HouseholdIncome) + labs(title="Multivitamin Usage by Household Income")
```
There are no brackets in which people are more likely to take a vitamin than not;but is there a signficant difference between brackets?
```{r}
fit <- aov(HSCReactive~HouseholdIncome, data=NHANESresearchreduced)
summary(fit)
#Yes, positive significant correlation. ad hoc test?

pairwise <- TukeyHSD(fit)
pairwise

IncomeHSCRPplot <- as.data.frame(pairwise[[1]])
IncomeHSCRPplot$conditions <- row.names(IncomeHSCRPplot)

ggplot(data=IncomeHSCRPplot, aes(x=conditions, y=diff)) +
  geom_point(size=2, color="coral") +
  geom_errorbar(aes(ymin=lwr, ymax=upr, width=.2)) +
  geom_hline(yintercept=0, color="coral", linetype="dashed") +
  labs(y="Differences in mean levels", x="",
       title="95% family-wise confidence levels") +
  theme_bw() + theme(axis.text.x = element_text(angle=45, hjust=1, size=10))
  coord_flip() 
```


```{r}
bmicat_vitamin<-xtabs(~Multivitamin + BMIcat, data=NHANESresearch)
prop.table(bmicat_vitamin)*100
NHANESresearch %>% 
  drop_na(Multivitamin) %>%
  ggplot(aes(x=Multivitamin)) + geom_bar(postition="dodge", color="coral", fill="ivory") + facet_wrap(~Gender) 


#Data %>% drop_na %>%  ggplot(aes(x=cat, y=cont)) + geom_boxplot(color="", fill="") + facet_wrap(~cont2) + labs(title="", x="", y="")
```

Is there a signficicant difference between individuals who do and do not take multivitamins?
T-test:
```{r}
NHANESresearch %>% 
  drop_na(Multivitamin) %>% 
  drop_na(HSCReactive) %>% 
  group_by(Multivitamin) %>% 
  summarize(mean(HSCReactive), sd(HSCReactive), count=n())

var.test(NHANESresearch$HSCReactive~NHANESresearch$Multivitamin, alternative="two.sided")

t.test(HSCReactive~Multivitamin, var.equal=FALSE, data=NHANESresearch %>%  drop_na(Multivitamin))

NHANESresearch %>% 
  drop_na(Multivitamin) %>% 
  drop_na(HSCReactive) %>% 
  ggplot(aes(x=Multivitamin, y=HSCReactive)) +
  geom_boxplot(color="coral", fill="ivory", width=.65) + 
  labs(title="Multivitamin Use and HSCRP", x="Daily Multivitamin Use", y="Mean High-Sensitivity C-Reactive Protein level") + theme_light()

#Filtered for outliers
NHANESresearch %>% 
  drop_na(Multivitamin) %>% 
  drop_na(HSCReactive) %>% 
  filter(HSCReactive<6.67) %>% 
  ggplot(aes(x=Multivitamin, y=HSCReactive)) +
  geom_boxplot(color="coral", fill="ivory", width=.65) + 
  labs(title="Multivitamin Use and HSCRP", x="Daily Multivitamin Use", y="Mean High-Sensitivity C-Reactive Protein level") + theme_light()
```
There is a statistically significant difference in levels of High-Sensitvity C-reactive proteins between people who take supplements and people who do not!!!　(^_^) Surprisingly, the mean HSCRP levels is higher in people who do than in people who don't.

Are there significant differences is BMI category between people who do and don't take multivitamins? 
```{r}
#table_name <- xtabs(~ CatVar1 + CatVar2, data=)
#chisq.test(table_name)

BMIcatHSCRP <- xtabs(~BMIcat + Multivitamin, data=NHANESresearch)
BMIcatHSCRP
prop.table(BMIcatHSCRP)*100
chisq.test(BMIcatHSCRP)

```
Yes! There is a significant difference between BMI groups in their multivtamin usage.


Correlation between BMI and HSCRP? 
```{r}
NHANESresearch %>% 
  drop_na(HSCReactive) %>% 
  group_by(BMIcat) %>% 
  drop_na(BMIcat) %>% 
  summarize(mean=mean(HSCReactive), sd=sd(HSCReactive), count=n())
            
            
NHANESresearch %>% 
  drop_na(HSCReactive) %>% 
  drop_na(BMIcat) %>% 
  dunn.test(x=NHANESresearch$HSCReactive, g=NHANESresearch$BMIcat, kw=TRUE)


#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(NHANESresearch$HSCReactive,NHANESresearch$BMI, method= "pearson")
#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()
ggplot(aes(x=HSCReactive, y=BMI), data=NHANESresearch) + geom_point(color="coral", alpha=.5) + geom_smooth(method=lm, color="wheat4") + labs(title= "Correlation of BMI to HSCRP levels", x="HSCRP (mg/L)", y="BMI")


```
There is a slight but significant positive correlation between BMI and HSCRP levels.


ANOVA and graph for BMI cat & HSCR
```{r}
bmihscfit <- aov(HSCReactive~BMIcat, data=NHANESresearch)
summary(bmihscfit)

BMIcatHSCRPplot <- NHANESresearch %>% 
  group_by(BMIcat) %>% 
  drop_na(BMIcat) %>% 
  drop_na(HSCReactive) %>% 
  summarize(n=n(),
            mean= mean(HSCReactive),
            sd=sd(HSCReactive),
            ci=qt(0.975, df=n-1)*sd/sqrt(n))
ggplot(BMIcatHSCRPplot, aes(x=BMIcat, y=mean, group= 1)) +
  geom_point(size=1, color="red", width=1.5) + 
  geom_errorbar(aes(ymin=mean-ci,
                    ymax=mean+ci),
                width=.1) +
  theme_bw() +
  labs(x="BMI Category",
       y="HSCRP",
       title="BMI Category and HSCRP with 95% CI")
```
P value= 0; There is a correlation between BMIcat and 

For odds test + chi squared, factor HSCRP into a dichotomous variable; identify healthy + unhealthy range?
Run odds ratio: Higher odds of High HSCRP by taking/not taking vitamins?
```{r}
multicrpodds <-table(NHANESresearch$Multivitamin, NHANESresearch$HSCRPlevel)
multicrpodds

oddsratio(multicrpodds)
#ggplot(data=database) + geom_mosaic(aes(x=product(IndVar), fill=(DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")


ggplot(data=NHANESresearch) + geom_mosaic(aes(x=product(Multivitamin), fill=HSCRPlevel), color="ivory", na.rm=TRUE, show.legend=FALSE) + labs(title="HSCRP Level by Multivitamin Usage", x= "Multivitamin", y= "HSCRP (mg/L)") + scale_fill_manual(values=c("Healthy"= "sandybrown", "High"="orangered"))
``` 
Yes- 1.13 Odds ratio- significant!! But it's not what I thought- People who take their multivitamins are more likely to have higher levels of HSCRP.
